---
title: "Normalitycheck"
author: "Chang Lu"
date: "2025-02-15"
output: pdf_document
---

```{r}
library(readxl)
library(dplyr)
library(nortest)
library(ggplot2)
```

### Data clean

```{r}
df <- read_excel("MAMS and Dental Combined for Analysis.xlsx", sheet = "Sheet1")
head(df)
summary(df)
```

```{r}
df_students <- df[2:32, ]

head(df_students)
# Define the new column names for the questions
question_names <- paste0("Q", 1:20)  # Generates Q1, Q2, ..., Q20

# Rename only the question columns
colnames(df_students)[2:21] <- question_names

head(df_students)
colnames(df_students)
```

```{r}
# Convert value format to numeric
df_students$`Total Score` <- as.numeric(df_students$`Total Score`)
df_students$`Quiz Time` <- as.numeric(df_students$`Quiz Time`)
df_students$Accomodations <- as.numeric(df_students$Accomodations)
df_students$Sex <- as.numeric(df_students$Sex)
df_students$`Race/Ethnicity` <- as.numeric(df_students$`Race/Ethnicity`)
df_students$`English Proficiency` <- as.numeric(df_students$`English Proficiency`)
df_students$`Born USA` <- as.numeric(df_students$`Born USA`)
df_students$`Home Language` <- as.numeric(df_students$`Home Language`)
df_students$`Age arrive USA`<- as.numeric(df_students$`Age arrive USA`)


# Convert Quiz Time from seconds to minutes
df_students$`Quiz Time Minutes` <- df_students$`Quiz Time` / 60 # for later test
# Create Quiz Time Bins
df_students$`Quiz Time Group` <- cut(df_students$`Quiz Time Minutes`, 
                                   breaks = c(0, 10, 20, Inf), 
                                   labels = c("0-10 min", "10-20 min", "20+ min"),
                                   include.lowest = TRUE)

# Check new groups
table(df_students$`Quiz Time Group`)


# Verify changes
str(df_students)
```
# Quiz_Time = check_normality_by_group(df_students, "Quiz Time Minutes","Total Score"),
# Accomodations = check_normality_by_group(df_students, "Accomodations","Total Score"),
# Age_Arrive_USA = check_normality_by_group(df_students, "Age arrive USA","Total Score")


### Shapiko-Wilk Test

```{r}
# function for normality check using the Shapiro-Wilk test
# if the result of p-value >0.05, the data is normal
check_normality_by_group <- function(data, group_col, test_col) {
  data %>%
    group_by(!!sym(group_col)) %>%
    summarise(
      Shapiro_Statistic = shapiro.test(na.omit(!!sym(test_col)))$statistic,
      P_Value = shapiro.test(na.omit(!!sym(test_col)))$p.value
    ) %>%
    arrange(P_Value)  
}
```

```{r}
# Check normality for Total Score by demographic group
normality_total_score <- list(
  Sex = check_normality_by_group(df_students, "Sex", "Total Score"),
  Race_Ethnicity = check_normality_by_group(df_students, "Race/Ethnicity", "Total Score"),
  English_Proficiency = check_normality_by_group(df_students, "English Proficiency", "Total Score"),
  Born_USA = check_normality_by_group(df_students, "Born USA", "Total Score"),
  Home_Language = check_normality_by_group(df_students, "Home Language", "Total Score")
)
```

```{r}
# Print results
print("Normality Test for Total Score by Group")
print(normality_total_score)
```
Both groups p > 0.05, we use t-test. At least one group p <= 0.05, we use Mann-Whitney U test (nonparametric).

Sex: MWU
Race: T-test
English: t-test
BornUSA: MWU
homeLanguage: t-test



```{r}

# Function to create a Q-Q plot for a given group
plot_qq <- function(data, group_col, test_col) {
  ggplot(data, aes(sample = !!sym(test_col))) +
    stat_qq() +
    stat_qq_line() +
    facet_wrap(~get(group_col)) +
    ggtitle(paste("Q-Q Plot for", test_col, "by", group_col)) +
    theme_minimal()
}

# Function to create a histogram with density plot for a given group
plot_histogram <- function(data, group_col, test_col) {
  ggplot(data, aes(x = !!sym(test_col))) +
    geom_histogram(aes(y = ..density..), bins = 20, fill = "lightblue", color = "black") +
    geom_density(color = "red", linewidth = 1) +
    facet_wrap(~get(group_col)) +
    ggtitle(paste("Histogram for", test_col, "by", group_col)) +
    theme_minimal()
}

# Generate normality plots for Total Score by different demographic groups
plot_qq(df_students, "Sex", "Total Score")
plot_histogram(df_students, "Sex", "Total Score")

plot_qq(df_students, "Race/Ethnicity", "Total Score")
plot_histogram(df_students, "Race/Ethnicity", "Total Score")

plot_qq(df_students, "English Proficiency", "Total Score")
plot_histogram(df_students, "English Proficiency", "Total Score")

plot_qq(df_students, "Born USA", "Total Score")
plot_histogram(df_students, "Born USA", "Total Score")

plot_qq(df_students, "Home Language", "Total Score")
plot_histogram(df_students, "Home Language", "Total Score")

plot_qq(df_students, "Accomodations", "Total Score")
plot_histogram(df_students, "Accomodations", "Total Score")

plot_qq(df_students, "Quiz Time Group", "Total Score")
plot_histogram(df_students, "Quiz Time Group", "Total Score")

plot_qq(df_students, "Age arrive USA", "Total Score")
plot_histogram(df_students, "Age arrive USA", "Total Score")
```

### Kolmogorov-Smirnov Test

```{r}
# function using the Kolmogorov-Smirnov Test
check_ks_normality_by_group <- function(data, group_col, test_col) {
  data %>%
    group_by(!!sym(group_col)) %>%
    summarise(
      Sample_Size = n(),
      KS_Statistic = ifelse(Sample_Size >= 3, 
                            ks.test(na.omit(!!sym(test_col)), "pnorm", 
                                    mean = mean(na.omit(!!sym(test_col))), 
                                    sd = sd(na.omit(!!sym(test_col))))$statistic, 
                            NA),
      P_Value = ifelse(Sample_Size >= 3, 
                       ks.test(na.omit(!!sym(test_col)), "pnorm", 
                               mean = mean(na.omit(!!sym(test_col))), 
                               sd = sd(na.omit(!!sym(test_col))))$p.value, 
                       NA)
    ) %>%
    arrange(P_Value)
}
```

```{r}
# Apply the KS test to different groups
ks_normality_accomodations <- check_ks_normality_by_group(df_students, "Accomodations", "Total Score")

print("Kolmogorov-Smirnov Normality Test for Total Score by Quiz Time Group")
print(ks_normality_accomodations)
```

# Lilliefors Test

```{r}
#  Function using Lilliefors Test
check_lilliefors_normality_by_group <- function(data, group_col, test_col) {
  data %>%
    group_by(!!sym(group_col)) %>%
    summarise(
      Sample_Size = n(),
      Lilliefors_Statistic = ifelse(Sample_Size >= 3, 
                                    lillie.test(na.omit(!!sym(test_col)))$statistic, 
                                    NA),
      P_Value = ifelse(Sample_Size >= 3, 
                       lillie.test(na.omit(!!sym(test_col)))$p.value, 
                       NA)
    ) %>%
    arrange(P_Value)
}
```

```{r}
lilliefors_normality_accomodations <- check_lilliefors_normality_by_group(df_students, "Accomodations", "Total Score")
lilliefors_normality_quiz_time <- check_lilliefors_normality_by_group(df_students, "Quiz Time Group", "Total Score")
lilliefors_normality_Age_Arrive_USA <- check_lilliefors_normality_by_group(df_students, "Age arrive USA", "Total Score")


print("Lilliefors Normality Test for Total Score by different Group")
print(lilliefors_normality_accomodations)
print(lilliefors_normality_quiz_time)
print(lilliefors_normality_Age_Arrive_USA)
```

I suggest using MWU test for accomodations and Age arrive USA, for one of the group sample size is too small(< 3), fails to using normality tests.

The quiz time group(categorized by 10 minutes) is suitable to use normality test.



### Conclusion: 

#### T-test: 

1. quiz time(categorized by 10 minutes)

2. English Proficiency

3. Race

4. Home Language

#### Mann-Whitney U Test

1. Accomodations

2. Age Arrive USA

3. Sex

4. Born in USA